---
title: "01_setup"
author: "Isabel García Valdivia"
date: "4/25/2018"
output: html_document
---
# PART I: Data Set-Up and Cleaning.
## Setup environment
I am using a R project, thus my working directory is already opened when I open my directory.  I have provided code for the manual set-up instructions below.
```{r, eval = F}
# remove all objects
rm(list=ls())
#Set and check directory (if you are not using a R project) - I am using a R Project, thus my directory is already set 
#Example of setting directory
#setwd(dir="/Users/isabelgarcia/Desktop/ps239T-final-project/Data")
#getwd()
```

## Load Libraries/Packages
```{r, eval=FALSE}
library(readstata13) #reads .dta Stata files constructed after Stata 12
library(dplyr)    #cleaning data
library(tidyr)    #cleaning data
```


## Load Data Set
The data set is from IPUMS Current Population Survey (CPS), particularly the data from the March Supplement from 2017 (we will start with 2017 data and use machine learning for 2016).
```{r, eval=FALSE}
# Read Stata file after Version 13 (see https://cran.r-project.org/web/packages/readstata13/README.html)
cps <- read.dta13("/Users/isabelgarcia/Dropbox/IPUMS_CPS/CPS_Data_2017/cps17_00019.dta") #CPS 2017 download: cps_00019
names(cps)  #gives you the names of all variables in data.frame
dim(cps) #check columns and rows

#Convert the variable names into lower case
colnames(cps)<-tolower(colnames(cps))
```

##Select & CleabVariables of Interest
The data contains 50 variables and more than 55,000 observations. We began by cleaning the data.

### Clean Variables
#### (1) Year
This variable was kept "as is".
```{r, eval=FALSE}
class(cps$year) #integer
table(cps$year)

cps$year_num <- as.numeric(cps$year) #convert to numeric
class(cps$year_num)
table(cps$year_num)
```

#### (2) Citizenship
This variable was converted to character and logical vectors. This includes the variable for individual person citizenship (named "citizen_chr" and "citizen_log") and spouse citizenship (named "citizensp_chr" and "citizensp_log").
```{r, eval=FALSE}
#Individual person
# (1) Factor
class(cps$citizen) #factor
table(cps$citizen) #need to recode
cps$citizen_factor <- cps$citizen #create a factor variable
levels(cps$citizen_factor) #Check levels and replace labels
levels(cps$citizen_factor)[5] <- NA
levels(cps$citizen_factor)[1] <- "us-born"
table(cps$citizen_factor) #check

# (2) Convert to character
cps$citizen_chr <- as.character(cps$citizen_factor)
table(cps$citizen_chr)

# (3) Convert to logical vector (2-level)
cps$citizen_log <- NA
cps$citizen_log[cps$citizen_chr == "us-born"] <- TRUE
cps$citizen_log[cps$citizen_chr == "born abroad of american parents"] <- TRUE
cps$citizen_log[cps$citizen_chr == "naturalized citizen"] <- TRUE
cps$citizen_log[cps$citizen_chr == "not a citizen"] <- FALSE
class(cps$citizen_log) #check
table(cps$citizen_log)
#TRUE = citizen
#FALSE = not citizen

#------------------------------------------------------------------#

#Spouse of person
# (1) Factor
class(cps$citizen_sp) #factor
table(cps$citizen_sp) #need to recode
cps$citizensp_factor <- cps$citizen_sp #create a factor variable
levels(cps$citizensp_factor) #Check levels and replace labels
levels(cps$citizensp_factor)[5] <- NA
levels(cps$citizensp_factor)[1] <- "us-born"
table(cps$citizensp_factor) #check

# (2) Convert to character
cps$citizensp_chr <- as.character(cps$citizensp_factor)
table(cps$citizensp_chr)

# (3) Convert to logical vector (2-level)
cps$citizensp_log <- NA
cps$citizensp_log[cps$citizensp_chr == "us-born"] <- TRUE
cps$citizensp_log[cps$citizensp_chr == "born abroad of american parents"] <- TRUE
cps$citizensp_log[cps$citizensp_chr == "naturalized citizen"] <- TRUE
cps$citizensp_log[cps$citizensp_chr == "not a citizen"] <- FALSE
class(cps$citizensp_log) #check
table(cps$citizensp_log)
#TRUE = citizen
#FALSE = not citizen
```

#### (3) Hispanic (ethnicity)
This variable was converted to character and logical vectors.
```{r, eval=FALSE}
# (1) Factor
class(cps$hispan)  #factor
table(cps$hispan)  #check
cps$hispan_factor <- cps$hispan
levels(cps$hispan_factor)
levels(cps$hispan_factor)[17] <- NA #N/A
levels(cps$hispan_factor)[16] <- NA #do not know
levels(cps$hispan_factor)[12:15] <- "central american"
levels(cps$hispan_factor)[8:11] <- "other latino"
levels(cps$hispan_factor)[2:7] <- "mexican"
levels(cps$hispan_factor)[1] <- "not latino"
table(cps$hispan_factor) #check

#(2) Convert to character vector (4-level)
cps$latino_chr <- as.character(cps$hispan_factor)
class(cps$latino_chr) #check 
table(cps$latino_chr) #check

#(3) Convert to logical vector (2-level)
cps$latino_log <- NA
cps$latino_log[cps$latino_chr == "central american"] <- TRUE
cps$latino_log[cps$latino_chr == "mexican"] <- TRUE
cps$latino_log[cps$latino_chr == "other latino"] <- TRUE
cps$latino_log[cps$latino_chr == "not latino"] <- FALSE
class(cps$latino_log) #check
table(cps$latino_log)
#TRUE = citizen
#FALSE = not citizen
```

#### (4) Race
This variable was converted to character vector and Hispanic/Latinx was added to it.
```{r, eval=FALSE}
# (1) Factor
class(cps$race) #factor
table(cps$race)
cps$race_factor <- cps$race #create factor with my edits
levels(cps$race_factor) #check the levels or the unique code for each
levels(cps$race_factor)[29] <- NA
levels(cps$race_factor)[7:28] <- "other"
levels(cps$race_factor)[4:6] <- "asian, pacific islander"
levels(cps$race_factor)[3] <- "other"
levels(cps$race_factor)[2] <- "black"
levels(cps$race_factor)[1] <- "white"
table(cps$race_factor)

#(2) Convert to character vector (4-level)
cps$race_chr <- as.character(cps$race_factor)
table(cps$race_chr) #inspect
class(cps$race_chr) #character

#(3) Overwrite race to add latino if latino is true
cps$race_chr[cps$latino_log==T]<-"latino"
table(cps$race_chr) #Inspect (again)
```

#### (5) Education 
This variable was converted to character and factor vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$educ)  #factor 
table(cps$educ)
cps$educ_factor <- cps$educ
levels(cps$educ_factor)
levels(cps$educ_factor)[36] <- NA
levels(cps$educ_factor)[33:35] <- "master's, professional, or doctoral"
levels(cps$educ_factor)[28:32] <- "bachelor's degree"
levels(cps$educ_factor)[25:27] <- "associate's degree"
levels(cps$educ_factor)[22:24] <- "some College"
levels(cps$educ_factor)[21] <- "high school diploma"
levels(cps$educ_factor)[3:20] <- "less than high school"
levels(cps$educ_factor )[2] <- NA #need to check this - it says it is left blank or niu (who are they)
levels(cps$educ_factor)[1] <- NA
table(cps$educ_factor)

# (2) Convert to character (6-options)
cps$educ_chr <- as.character(cps$educ_factor)
class(cps$educ_chr)
table(cps$educ_chr)
```

#### (6) Gender
This variable was converted to character and logical vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$sex)   #factor
table(cps$sex)   #Female = 2, Male = 1, niu
levels(cps$sex)
cps$female_factor <- cps$sex
levels(cps$female_factor)[3] <- NA
levels(cps$female_factor) #check again
table(cps$female_factor) #check

#(1) Convert to character
cps$female_chr<-as.character(cps$female_factor)
class(cps$female_chr)
table(cps$female_chr)

#(2) Convert to logical
cps$female_log <- NA
cps$female_log[cps$female_chr == "female"] <- TRUE
cps$female_log[cps$female_chr == "male"] <- FALSE
class(cps$female_log)
table(cps$female_log)
```

#### (7) Age
This variable was converted to character vector and we also kept it as an integer vector.
```{r, eval=FALSE}
#(1) Factor
class(cps$age) #factor
table(cps$age) #ranges from 0 - 85
levels(cps$age) #some adnormal things; "under 1 year"; 
cps$age_factor <- cps$age
table(cps$age_factor)
levels(cps$age_factor)[1] <- "0" #convert the individuals that are less than 1 to 0
summary(cps$age_factor)

cps$age_factor2 <- cps$age_factor #create another one because you will make into categories
levels(cps$age_factor2)  #need to start at 51 becaue of way factor is set up
  levels(cps$age_factor2)[66:100] <- "65 and over"
  levels(cps$age_factor2)[61:65] <- "60 to 64 years"
  levels(cps$age_factor2)[56:60] <- "55 to 59 years"
  levels(cps$age_factor2)[51:55] <- "50 to 54 years"
  levels(cps$age_factor2)[1:50] <-  NA #"under 50"
  
table(cps$age_factor2)

#(2) Convert to character
cps$age_chr <- as.character(cps$age_factor2)
table(cps$age_chr)

#(3) Convert to numeric
cps$age_num<- as.numeric(cps$age_factor) - 1 
#subract to include the children under 1 (coded as 0) because if not it adds 1 to each of the ages 
table(cps$age_num)
table(cps$age_num >= 50) #58486 equal or greater 50 years old

#(4) Convert to integer
cps$age_int<- as.integer(cps$age_factor) -1 
table(cps$age_int)
table(cps$age_int >= 50)
```

#### (8) Year Migrated
This variable was converted to character and factor vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$yrimmig) #factor
table(cps$yrimmig) #Need to get rid of 0 = not in universe/not migrants
levels(cps$yrimmig)

cps$yrimmig_factor <- cps$yrimmig
levels(cps$yrimmig_factor)
  levels(cps$yrimmig_factor)[33:39] <- "2010s"
  levels(cps$yrimmig_factor)[22:32] <- "2000s"
  levels(cps$yrimmig_factor)[13:21] <- "1990s"
  levels(cps$yrimmig_factor)[8:12] <- "1980s"
  levels(cps$yrimmig_factor)[2:7] <- "Before 1980"
  levels(cps$yrimmig_factor)[1] <- NA #not in universe (niu)
table(cps$yrimmig_factor)  
```

#### (9) Marital Status
This variable was converted to character and factor vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$marst)  #factor
table(cps$marst)
levels(cps$marst)

cps$marst_factor <- cps$marst
levels(cps$marst_factor)[8] <- NA 
  levels(cps$marst_factor)[7] <- "widowed or divorced"
  levels(cps$marst_factor)[6] <- "never married/single"
  levels(cps$marst_factor)[5] <- "widowed or divorced"
  levels(cps$marst_factor)[4] <- "widowed or divorced"
  levels(cps$marst_factor)[3] <- "separated"
  levels(cps$marst_factor)[1:2] <- "married"
table(cps$marst_factor)

#(2) Convert to character
cps$marst_chr<-as.character(cps$marst_factor)
class(cps$marst_chr)
table(cps$marst_chr)
```

#### (10) Nativity
This variable was converted to a character vector.
```{r, eval=FALSE}
class(cps$nativity) #factor
table(cps$nativity)
levels(cps$nativity)

#(1) Convert to character
cps$nat_chr<-as.character(cps$nativity)
class(cps$nat_chr)
table(cps$nat_chr)
```

#### (11) Birthplace (foreign-born)
This variable was converted to character and factor vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$bpl)  #factor
table(cps$bpl)

cps$bpl_factor <- cps$bpl
levels(cps$bpl_factor)
  levels(cps$bpl_factor)[173] <- NA
  levels(cps$bpl_factor)[172] <- "other" 
  levels(cps$bpl_factor)[164:171] <- "australia/oceania" 
  levels(cps$bpl_factor)[136:163] <- "africa"
  levels(cps$bpl_factor)[93:135] <- "asia" 
  levels(cps$bpl_factor)[92] <- "europe"
  levels(cps$bpl_factor)[91] <- "asia"  #ussr
  levels(cps$bpl_factor)[47:90] <- "europe"
  levels(cps$bpl_factor)[34:46] <- "south america"
  levels(cps$bpl_factor)[21:33] <- "carribbean"
  levels(cps$bpl_factor)[20] <- "cuba" 
  levels(cps$bpl_factor)[12:19] <-"central america"
  levels(cps$bpl_factor)[11] <- "mexico"
  levels(cps$bpl_factor)[10] <- "north america"
  levels(cps$bpl_factor)[9] <- "carribbean"
  levels(cps$bpl_factor)[8] <- "canada"
  levels(cps$bpl_factor)[2:7] <- "united states, territories" 
  levels(cps$bpl_factor)[1] <- "united states" 
table(cps$bpl_factor)

#(2) Convert to character
cps$bpl_chr<-as.character(cps$bpl_factor) #making a copy so I can recode the numerics
table(cps$bpl_chr)
```


#### (12) Occupation
This variable was converted to character and factor vectors. *Note:* The only variables used were those that we could theoretically argue required legal status.
```{r, eval=FALSE}
#(1) Factor
class(cps$occ2010)  #factor
levels(cps$occ2010)

cps$occ_factor <- cps$occ2010
  levels(cps$occ_factor)[454:457] <- "military"
  levels(cps$occ_factor)[441] <- "transportation inspectors" 
  levels(cps$occ_factor)[434:438] <- "transportation: inspectors, bus/ambulance drivers, flight attendants"
  levels(cps$occ_factor)[426:429] <- "transportation: inspectors, bus/ambulance drivers, flight attendants"
  levels(cps$occ_factor)[413] <- "medical"
  levels(cps$occ_factor)[261:263] <- "post office" #need to be citizen, lpr 
  levels(cps$occ_factor)[164:167] <- "law enforcement" 
  levels(cps$occ_factor)[129:163] <- "medical"
  levels(cps$occ_factor)[102:112] <- "teachers with licenses"
  levels(cps$occ_factor)[99:101] <- "lawyers and legal" 
levels(cps$occ_factor)
#here you united certain jobs that theoretically require legal status
  
#(2) Convert to character and take out these labels in a logical T/F
cps$occ_chr <- as.character(cps$occ_factor)
class(cps$occ_chr)

cps$occlegal_log <- F   #False if we cannot make a theoretical arguement that it requires legal status; True if we are sure it requires legal status 15277 
cps$occlegal_log[cps$occ_chr == "military"] <- T
cps$occlegal_log[cps$occ_chr == "transportation inspectors"] <- T
cps$occlegal_log[cps$occ_chr == "transportation: inspectors, bus/ambulance drivers, flight attendants"] <- T
cps$occlegal_log[cps$occ_chr == "medical"] <- T
cps$occlegal_log[cps$occ_chr == "post office"] <- T
cps$occlegal_log[cps$occ_chr == "law enforcement"] <- T
cps$occlegal_log[cps$occ_chr == "teachers with licenses"] <- T
cps$occlegal_log[cps$occ_chr == "lawyers and legal"] <- T
table(cps$occlegal_log)

```

#### (13) Class of Worker
This variable was kept as a factor vector.
```{r, eval=FALSE}
#(1) Factor
class(cps$classwkr) #factor
table(cps$classwkr)

cps$classwkr_factor <- cps$classwkr
levels(cps$classwkr_factor)
  levels(cps$classwkr_factor)[15] <- NA
  levels(cps$classwkr_factor)[14] <- "unpaid family worker"
  levels(cps$classwkr_factor)[12:13] <- "local, state, federal government"
  levels(cps$classwkr_factor)[11] <- "military"
  levels(cps$classwkr_factor)[10] <- "local, state, federal government"
  levels(cps$classwkr_factor)[9] <- "local, state, federal government"
  levels(cps$classwkr_factor)[5:8] <- "wages"
  levels(cps$classwkr_factor)[2:4] <- "self-employed"
  levels(cps$classwkr_factor)[1] <- NA
table(cps$classwkr_factor)
```

#### (14) Covered by Medicaid
This variable was converted to character, factor, and logical vectors.
```{r, eval=FALSE}
#(1) Factor
class(cps$caid)
table(cps$caid)

#(2) Convert to character
cps$caid_chr<-as.character(cps$caid)
class(cps$caid_chr)
table(cps$caid_chr)

#(3) Convert to logical
cps$caid_log <- ifelse(cps$caid == "yes", TRUE, FALSE) #conver to logical - if yes than equal to TRUE
class(cps$caid_log)
table(cps$caid_log)
```

#### (15) Got WIC (WIC, the Special Supplemental Nutrition Program for Women, Infants and
Children) & Food stamps
These variables were converted to character, factor, and logical vectors.
```{r, eval=FALSE}
table(cps$gotwic)
#0; NIU ~ men
#1: No - 48233
#2: Yes - 2241
cps$gotwic_factor <- cps$gotwic
levels(cps$gotwic_factor)
  levels(cps$gotwic_factor)[1] <- NA #niu - not in universe
  table(cps$gotwic_factor)

#(1) Convert to logical
cps$gotwic_log <- ifelse(cps$gotwic == "yes", TRUE, FALSE) #note that NIU is under false (including men)
class(cps$gotwic_log) #check
table(cps$gotwic_log)

#(2) Convert to character
cps$gotwic_chr <- as.character(cps$gotwic_factor)
class(cps$gotwic_chr) #check
table(cps$gotwic_chr)

#Food Stamps
#(1) Factor
class(cps$foodstmp) #factor
table(cps$foodstmp)
levels(cps$foodstmp)
  levels(cps$foodstmp)[1] <- NA #niu = 0
    table(cps$foodstmp)
#(2) Convert to logical
cps$foodstmp_log <- ifelse(cps$foodstmp == "yes", TRUE, FALSE)
table(cps$foodstmp_log)

```

#### (16) Covered by Medicaid and/or Medicare 
These variables were converted to character, factor, and logical vectors.
```{r, eval=FALSE}
#Medicaid Coverage
class(cps$himcaid)  #factor
table(cps$himcaid)

#(1) Convert to logical
cps$himcaid_log <- ifelse(cps$himcaid == "yes", TRUE, FALSE)
class(cps$himcaid_log)
table(cps$himcaid_log)  #check

#------------------------------------------------------------------#

#Medicare Coverage
class(cps$himcare)  #factor
table(cps$himcare)
levels(cps$himcare)

cps$himcare_factor <- cps$himcare
  levels(cps$himcare_factor)
  levels(cps$himcare_factor)[1] <- NA
table(cps$himcare_factor)
  
#(1) Convert to logical
cps$himcare_log <- ifelse(cps$himcare == "yes", TRUE, FALSE)
class(cps$himcare_log)
table(cps$himcare_log)  #Check
```

#### (17) Receives Social Security benefits or SSI
These variables were converted to character, factor, and logical vectors.
```{r, eval=FALSE}
#Social Security benefits
class(cps$incss)  #integer
#table(cps$incss)
summary(cps$incss)
summary(cps$incss == 0)  #no income from issi
summary(cps$incss > 0 & cps$incss < 99999) #includes anyone that gets an amount from incss
summary(cps$incss == 99999) #not in universe

#(1) Convert to logical
cps$incss_log <- TRUE
cps$incss_log[cps$incss <= 0] <- FALSE
cps$incss_log[cps$incss == 99999] <- NA
table(cps$incss_log)  #check
#------------------------------------------------------------------#

#SSI
class(cps$incssi) #integer
summary(cps$incssi)
summary(cps$incssi == 99999) #not in universe
summary(cps$incssi == 0)  #do not recieve = 0
summary(cps$incssi > 0 & cps$incssi < 99999) #anyone that gets an amount from incssi
#(1) Convert to logical
cps$incssi_log <- TRUE
cps$incssi_log[cps$incssi <= 0] <- FALSE
cps$incssi_log[cps$incssi == 99999] <- NA
table(cps$incssi_log) #check
```

#### (18) Miltary Info
These variables were converted to logical vectors.
```{r, eval=FALSE}
#Military Insurance
class(cps$hichamp)  #factor
table(cps$hichamp)
#(1) Convert to logical
cps$hichamp_log <- ifelse(cps$hichamp == "yes", TRUE, FALSE)
table(cps$hichamp_log)

#------------------------------------------------------------------#

#Veteran
class(cps$vetstat)  #factor
table(cps$vetstat)

cps$vetstat_factor <- cps$vetstat
levels(cps$vetstat_factor) #check
  levels(cps$vetstat_factor)[4] <- NA   #unknown
    levels(cps$vetstat_factor)[3] <- "yes" 
    levels(cps$vetstat_factor)[2] <- "no"
    levels(cps$vetstat_factor)[1] <- NA
table(cps$vetstat_factor)

#(1) Convert to logical
cps$vetstat_log <- ifelse(cps$vetstat_factor =="yes", TRUE, FALSE)
table(cps$vetstat_log)  #check
```

#### (19) Person resides in public housing or receives rental subsidies (or their spouse does)
These variables are at the household-level, not on the person-level (that is why you need to include spouse in the language). These variables were converted to logical vectors.
```{r, eval=FALSE}
#Lives in public housing
#(1) Factor
class(cps$pubhous)  #factor
table(cps$pubhous)  #check

cps$pubhous_factor <- cps$pubhous
levels(cps$pubhous_factor)
levels(cps$pubhous_factor)[1] <- NA
table(cps$pubhous_factor)

#(2) Convert to logical
cps$pubhous_log <- ifelse(cps$pubhous_factor =="yes", TRUE, FALSE)
table(cps$pubhous_log)  #check

#Paying lower rent due to government subsidy
#(1) Factor
class(cps$rentsub)  #factor
table(cps$rentsub)  #check

cps$rentsub_factor <- cps$rentsub
levels(cps$rentsub_factor)
  levels(cps$rentsub_factor)[1] <- NA
table(cps$rentsub_factor)

#(2) Convert to logical
cps$rentsub_log <- ifelse(cps$rentsub_factor =="yes", TRUE, FALSE)
table(cps$rentsub_log)  #check
```

#### (20) Geographic Area
This variable was converted to character vector.
```{r, eval=FALSE}
#(1) Factor
class(cps$statefip) #factor
table(cps$statefip)

cps$statefip_factor <- cps$statefip
levels(cps$statefip_factor)
levels(cps$statefip_factor)[52:75] <- NA  #They have "0" and they are more like areas/not states
table(cps$statefip_factor)

#(2) Convert to character
cps$state_chr <- as.character(cps$statefip_factor)
class(cps$state_chr)
table(cps$state_chr)
```

#### (21) Mexican-origin Immigrants
This isolated the foreign-born immigrants born in Mexico.
```{r, eval=FALSE}
class(cps$bpl_chr)
table (cps$bpl_chr)

mexico_bpl <- cps$bpl_chr == "mexico"
table(mexico_bpl) #check
class(mexico_bpl) #logical
```

#### Variable of interest: Legal status
First, I created a logical variable (undoc_log) where TRUE means undocumented and FALSE identifies immigrants with legal status.
```{r, eval=FALSE}
#Has Legal Status
cps$undoc_log <- TRUE #58486 observations

cps$undoc_log[cps$citizen_log == T] <- F #Person-level Citizen
cps$undoc_log[cps$citizensp_log == T] <- F #Spouse Citizen
cps$undoc_log[cps$yrimmig_factor == "Before 1980"] <- F #Came before 1980

cps$undoc_log[cps$incss_log == T] <- F #SS benefits
cps$undoc_log[cps$incssi_log == T] <- F #SSI benefits
cps$undoc_log[cps$himcaid_log == T] <- F #Medicaid
cps$undoc_log[cps$himcare_log == T] <- F #Medicare

cps$undoc_log[cps$hichamp_log == T] <- F #Military health insurance
cps$undoc_log[cps$vetstat_log == T] <- F #Veteran
cps$undoc_log[cps$classwkr_factor == "military"] <- F #Same as above
cps$undoc_log[cps$classwkr_factor == "local, state, federal government"] <- F #Government Sector (federal, military, state, or local) #1284

cps$undoc_log[cps$pubhous_log == T] <- F #public housing
cps$undoc_log[cps$rentsub_log == T] <- F #subsidy
cps$undoc_log[cps$bpl == "cuba"] <- F #born in Cuba
cps$undoc_log[cps$occlegal_log == T] <- F   #Occupation requires licencing 

table(cps$undoc_log) #Check the variable
sum(cps$undoc_log*cps$asecwt) #Times the undoc variable with the individual_level cps weights
#1,958,698 people (over estimated because of all population there is an estimated 11.5 - 11.7 per literature - probably occupation)

#Add food stamps - now there seems to be an undercount - 1,825,436
cps$undoc_log[cps$foodstmp_log == T] <- F #if they get food stamps (at household-level) 1082
table(cps$undoc_log) #check - undercount
sum(cps$undoc_log*cps$asecwt) #Times the undoc variable with the individual_level cps weights
sum((cps$undoc_log == T)*cps$asecwt) #check another way
```

Now, I will reconstruct the citizen variable to incorporate undocumented status.
```{r, eval=FALSE}
table(cps$citizen_factor) #factor
levels(cps$citizen_factor)  #get the number of levels

cps$legalstat_factor <- cps$citizen_factor #create a new var
  levels(cps$legalstat_factor)
  levels(cps$legalstat_factor)[3:4] <- "legal immigrants"
  levels(cps$legalstat_factor)[1:2] <- "us-born and born abroad"
  table(cps$legalstat_factor)

#(1) Convert to character, add undocumented population, and afterwards rename non-citizen to legal non-citizen
cps$legalstat_chr <- as.character(cps$legalstat_factor) #convert to character 
cps$legalstat_chr[cps$undoc_log == T] <- "undocumented" #add undoc population - convert those without legal status
cps$legalstat_chr[cps$legalstat_chr == "not a citizen"] <- "legal immigrants" #change the names
table(cps$legalstat_chr)
```

### Weights
For most person-level analyses of the ASEC samples, apply the ASECWT variable. ASECWT gives the population represented by each individual in the sample.  The sample represents 111,962,818 (of 320,371,997 people ~U.S. population - according to full cps data for 2017) of individuals 50 years and older in 2017.
```{r, eval=FALSE}
sum(cps$asecwt)
```

#### Overall Calculations with Weights
```{r, eval=FALSE}
#check that weights are working correctly with undoc variable 
sum((cps$undoc_log==F)*cps$asecwt) #110,137,383
sum((cps$undoc_log==T)*cps$asecwt) #1,825,436
count <- 1825436 + 110137383
count #111962819

sum((cps$legalstat_chr == "legal immigrants" & cps$nat_chr == "foreign born")*cps$asecwt) #14,501,056
sum((cps$legalstat_chr == "undocumented" & cps$nat_chr == "foreign born")*cps$asecwt) #1,813,288 undocumented
sum((cps$legalstat_chr == "us-born and born abroad")*cps$asecwt) #95,595,635 includes born abroad

sum((cps$legalstat_chr == "legal immigrants" & mexico_bpl == T)*cps$asecwt) #3,017,767
sum((cps$legalstat_chr == "undocumented" & mexico_bpl == T)*cps$asecwt) #743,736 undocumented
sum((cps$legalstat_chr == "us-born and born abroad" & cps$latino_chr == "mexican")*cps$asecwt) #2,950,808 includes born abroad
```

#### Subset Data for Age Pyramids
```{r, eval=FALSE}
#Group all indviduals by legal status (includes nat. citizens and residents)
legalfor_df <- filter(cps, nat_chr == "foreign born" & legalstat_chr == "legal immigrants")
#Group all individuals by undocumented
undocfor_df <- filter(cps, nat_chr == "foreign born" & legalstat_chr == "undocumented")
#Group us-born 
usborn_df <- filter(cps, legalstat_chr == "us-born and born abroad")

###These graphs will focus on Mexican foreign-born by legal status
# Group all individuals by mexican_bpl and age 
usbornmex_df <- filter(cps, latino_chr == "mexican" & legalstat_chr == "us-born and born abroad")
# Group all individuals by mexican_bpl and age 
legalmex_df <- filter(cps, mexico_bpl == T & nat_chr == "foreign born" & legalstat_chr == "legal immigrants")  #create dataset with mexicans only - over 50 years old
# Group all individuals by mexican_bpl and undoc
undocmex_df <- filter(cps, mexico_bpl == T & nat_chr == "foreign born" & legalstat_chr == "undocumented")  #create dataset with mexicans only - over 50 years old
```

### Saving Files and Data Preparation for Machine Learning
```{r, eval=FALSE}
#Save the data as .csv
write.csv(cps, "cps17.csv")
#You need to make sure you save different years of the CPS accordingly - for example, cps16, cps17

#create another data set for cps17 machine learning
myvars <- c("year", "serial", "month", "cpsid", "asecflag", "asecwth", "statefip", "metro",
            "metarea", "county", "statecensus", "ownershp","pubhous", "rentsub", "foodstmp", "pernum", "cpsidp",
            "asecwt", "age", "sex", "race", "marst", "vetstat", "famsize", "nchild", "bpl", "yrimmig", "citizen", "mbpl",
            "fbpl", "nativity", "hispan", "empstat", "occ2010", "classwkr", "educ", "incss", "incssi", "himcaid",
            "himcare", "hichamp", "caid", "gotwic", "bpl_sp", "yrimmig_sp", "citizen_sp", "mbpl_sp", "fbpl_sp",
            "nativity_sp", "hispan_sp", "undoc_log")

cps17_mlearning <- cps[myvars] #make variables into new dataframe
write.csv(cps17_mlearning, "cps17_mlearning.csv") #save for machine learning - all original variables plus undoc_log

#Prepare 2016 data
cps16_mlearning <- read.dta13("/Users/isabelgarcia/Dropbox/IPUMS_CPS/CPS_Data_2017/cps16_00020.dta")
write.csv(cps16_mlearning, "cps16_mlearning.csv")
```