---
title: "02_visualizations"
author: "Isabel García Valdivia"
date: "4/25/2018"
output: html_document
---

#PART II. Data Tables and Plots.
Constructing data tables and plots with new legal status variable (legalstat) in (cps.dta) data set to look at the data numerically and visually.

## Load Libraries/Packages
```{r, eval=FALSE}
library(vcd)      #three-way crostabs
library(descr)    #weighted data and cross tabs
library(xtable)   #export to LaTeX
library(ggplot2)  #visuals
library(ggthemes) #changing visual themes
library(gridExtra) #packages to export plots
library(cowplot) #same as above
```

#### Tabulations of the Data: Older Adults by Legal Status and Age
```{r, eval=FALSE}
#Legal status and age
table1 <- crosstab(cps$age_chr, cps$legalstat_chr, #DV, IV
         weight = cps$asecwt, #weight
         digits = list(prop = 3, percent = 1), #Named list with number of digits after the decimal point for four categories of statistics: expected values, cell proportions, percentage and others statistics
         prop.c =  TRUE, #column percents
         xlab = "Legal Status", #x-axis label
         ylab = "Age", #y-axis label
         main = "Old Age by Legal Status", #overall title
         plot = F) #Logical: if TRUE (default), a mosaic plot is produced.
table1
xtable(table1) #export to LaTeX
```


####Boxplot: Older Adults by Legal Status and Age
```{r, eval=FALSE}
#Graph 1: boxplot legal status by age
boxplot1 <- ggplot(cps, aes(x= legalstat_chr, y= age_num, fill=legalstat_chr)) +
  geom_boxplot(varwidth = TRUE) +#boxplot #varwidth - changes according to the number of observations
  theme_classic() + #theme
  theme(plot.title = element_text(hjust = 0.5),
        legend.position='none',
        legend.title = element_blank()) + #Center title and move legend
  #labs(title = "Distribution of Old Age (50+ years old) in the U.S.",
       labs(y = "Age (years)") + #labels for title and y-axis
  scale_x_discrete(name= "Legal Status", #re-label the x-axis labels 
                  labels=c("legal immigrants","undocumented", "us-born")) 

boxplot1 #check graph
ggsave("g1_boxplot.png",
       boxplot1,
       dpi = 600)  #save the file as .png
```

###Graphs 1-3: Displays graphs of older adults (50+ years) by legal status and gender.
####Graph 1: Legal Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb1 <- as.data.frame(table(legalfor_df$age_chr, legalfor_df$female_log))
names(pop_tb1)[1] <- paste("Age")  #change the label name
names(pop_tb1)[2] <- paste("Female") #change the label name
sum(pop_tb1$Freq) #foreign-born and legal


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g1_legalfor <- ggplot(pop_tb1, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb1, Female == F), stat = "identity", mapping = aes(y = (Freq/8058)*100)) + #men
  geom_bar(data = subset(pop_tb1, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/8058)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('Legal Immigrants by Age and Gender (3.8 million)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within Legal Foreign-Born") 


g1_legalfor #check graph
ggsave("g1_legalimmigrants.png", 
       g1_legalfor,
       dpi = 600)  #save the file as .png
```

####Graph 2: Undocumented Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb2 <- as.data.frame(table(undocfor_df$age_chr, undocfor_df$female_log))
names(pop_tb2)[1] <- paste("Age")  #change the label name
names(pop_tb2)[2] <- paste("Female") #change the label name
sum(pop_tb2$Freq) #foreign-born and undoc


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g2_undocfor <- ggplot(pop_tb2, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb2, Female == F), stat = "identity", mapping = aes(y = (Freq/1075)*100)) + #men
  geom_bar(data = subset(pop_tb2, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/1075)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('Undocumented Immigrants by Age and Gender (1.8 million)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within Undocumented Foreign-Born") 


g2_undocfor #check graph
ggsave("g2_undocimmigrants.png", 
       g2_undocfor,
       dpi = 600)  #save the file as .png
```


####Graph 3: U.S. Born Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb3 <- as.data.frame(table(usborn_df$age_chr, usborn_df$female_log))
names(pop_tb3)[1] <- paste("Age")  #change the label name
names(pop_tb3)[2] <- paste("Female") #change the label name
sum(pop_tb3$Freq) #usborn


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g3_usborn <- ggplot(pop_tb3, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb3, Female == F), stat = "identity", mapping = aes(y = (Freq/49323)*100)) + #men
  geom_bar(data = subset(pop_tb3, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/49323)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('U.S. Born by Age and Gender (95.6 million)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within U.S. Born") 


g3_usborn#check graph
ggsave("g3_usborn.png", 
       g3_usborn,
       dpi = 600)  #save the file as .png

#Export Graphs 1-3 titled olderadultsinUS.png
olderadultsinUS <- plot_grid(g1_legalfor, g2_undocfor, g3_usborn, align = "v", nrow = 3, rel_heights = c(1/2, 1/2))
png(file = "olderadultsinUS.png", width = 1500, height = 1000, res = 250)
plot_grid(g1_legalfor, g2_undocfor, g3_usborn, align = "v", nrow = 3, rel_heights = c(1/2, 1/2))
dev.off() #end save
```

###Graphs 4-6: Displays graphs of Mexican older adults (50+ years) by legal status and gender.
####Graph 4: Mexican U.S. Born Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb4 <- as.data.frame(table(usbornmex_df$age_chr, usbornmex_df$female_log))
names(pop_tb4)[1] <- paste("Age")  #change the label name
names(pop_tb4)[2] <- paste("Female") #change the label name
sum(pop_tb4$Freq) #usborn mex


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g4_usbornmex <- ggplot(pop_tb4, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb4, Female == F), stat = "identity", mapping = aes(y = (Freq/2033)*100)) + #men
  geom_bar(data = subset(pop_tb4, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/2033)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('U.S. Born Mexicans by Age and Gender (2.9 million)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within U.S. Born Mexicans") 

g4_usbornmex #check graph
ggsave("g4_usbornmex.png", 
       g4_usbornmex,
       dpi = 600)  #save the file as .png
```

####Graph 5: Mexican Legal Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb5 <- as.data.frame(table(legalmex_df$age_chr, legalmex_df$female_log))
names(pop_tb5)[1] <- paste("Age")  #change the label name
names(pop_tb5)[2] <- paste("Female") #change the label name
sum(pop_tb5$Freq) #usborn mex


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g5_legalmex <- ggplot(pop_tb5, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb5, Female == F), stat = "identity", mapping = aes(y = (Freq/1955)*100)) + #men
  geom_bar(data = subset(pop_tb5, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/1955)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('Legal Mexican Immigrants by Age and Gender (3 million)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within Legal Mexican Immigrants") 

g5_legalmex #check graph
ggsave("g5_legalmex.png", 
       g5_legalmex,
       dpi = 600)  #save the file as .png
```

####Graph 6: Mexican Undocumented Immigrants by Age and Gender
```{r, eval=FALSE}
pop_tb6 <- as.data.frame(table(undocmex_df$age_chr, undocmex_df$female_log))
names(pop_tb6)[1] <- paste("Age")  #change the label name
names(pop_tb6)[2] <- paste("Female") #change the label name
sum(pop_tb6$Freq) #477 undoc


#Plot Age Range with respective frequencies and label with mexican_bpl by age
g6_undocmex <- ggplot(pop_tb6, aes(x = Age, y = Freq, fill = Female)) + 
  # Define each legal status in barplot distribution
  geom_bar(data = subset(pop_tb6, Female == F), stat = "identity", mapping = aes(y = (Freq/477)*100)) + #men
  geom_bar(data = subset(pop_tb6, Female == T), stat = "identity", position = "identity", mapping = aes(y = (-Freq/477)*100)) + #female #freq/total number of legal foreign-born *100 gives percent of legal pop 50 and over
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"), #Change colors
                    name="", labels=c("Men", "Women")) + #Change the names of legend labels for "fill"
  coord_flip() + # Turn to horizontal barplots
  #ggtitle('Undocumented Mexican Immigrants by Age and Gender (700 Thousands)') +
  #theme(plot.title = element_text(hjust = 0.5)) + #Center title
  theme_classic() +
  scale_y_continuous(breaks = seq(-30, 30, 5), # Set interval for breaks for y axis
                     limits = c(-30, 30), # Define/set the limits on scales
                     labels = paste0(as.character(c(seq(30, 0, -5), seq(5, 30, 5))))) + # Label the limits on scales
  xlab("Age (years)") + 
  ylab("% Within Undocumented Mexicans") 

g6_undocmex #check graph
ggsave("g6_undocmex.png", 
       g6_undocmex,
       dpi = 600)  #save the file as .png

#Export Graphs 4-6 titled olderadultsinUS.png
mexolderadultsinUS <- plot_grid(g5_legalmex, g6_undocmex, g4_usbornmex, align = "v", nrow = 3, rel_heights = c(1/2, 1/2))
png(file = "mexolderadultsinUS.png", width = 1500, height = 1000, res = 250)
plot_grid(g5_legalmex, g6_undocmex, g4_usbornmex, align = "v", nrow = 3, rel_heights = c(1/2, 1/2))
dev.off() #end save
```
